\Title

### 2023-02-19 Grimaldi-etal-BiolCybernetics.tex
new figures

This research was funded by the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{APROVIS3D}{http://aprovis3d.eu/}''). % Agence Nationale de la Recherche %HL and 

@@

% LP received funding 
and from A*Midex grant number AMX-21-RID-025
``\href{Polychronies}{https://laurentperrinet.github.io/grant/polychronies/}''. 
%HL received funding from a University of Montreal Artificial Intelligence scholarship. 
%
}
\newcommand{\Acknowledgments}{
Thanks to % Thomas Schatz, Rosa Cossart,
Hugo Ladret, Camille Besnainou and Jean-Nicolas Jérémie for useful discussions prior to the elaboration of this work. %Thanks to HL  for participating to Mustache November ©. %
\Funding %
%For the purpose of open access, the author has applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission. This research was funded, in whole or in part, by [Organisation name, Grant #]. 
A CC-BY public copyright license has been applied by the authors to the present document and will be applied to all subsequent versions up to the Author Accepted Manuscript arising from this submission, in accordance with the grant’s open access conditions. 
}
\newcommand{\DataAvailability}{
This works is made reproducible using the following tools. First the code reproducing all figures is available on~\href{ https://github.com/SpikeAI/pyTERtorch}{GitHub}. % todo: open-up the repo
Find also the associated~\href{https://www.zotero.org/groups/4776796/fastmotiondetection}{zotero group} used to gather relevant literature on the subject.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\jyear{2023}%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

@@

%\usepackage{lineno} %Adds line numbers to selected paragraphs with reference possible through the LaTeX \ref and \pageref cross reference mechanism.

@@

% Title.
% ------

@@

% Single address.

@@

%%==================================%%
%% sample for unstructured abstract %%
%%==================================%%

\abstract{
\Abstract
}

%\ninept
%
%\abstract{Abstract}
%
\keywords{\Keywords}
%
\maketitle
%%%%%%%%% BODY TEXT
\section{Introduction}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable property of being able to react at any time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation both use. Artificial convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as tensors, storing visual information densely across the visual topography, with different properties represented in different channels. These networks are known to mimic many properties of biological systems, such that each can be attributed a ``Brain Score''~\citep{schrimpf_brain-score_2020}, yet this score takes no account of inference speed or energy consumption. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. However, they do not take advantage of the dynamics inherent in the way we perceive our natural environment. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. The third generation of artificial neural networks, known as spiking neural networks (SNNs), incorporates this temporal dimension into the way they perform their computations. They are of interest for computational neuroscience because they provide a better model of the biological neuron. Some approaches have developed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses accurate timing of spikes attempts to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been recently extended. This method uses a surrogate gradient and is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of firing-rate based networks. The question of the advantage of using spikes in machine learning and computer vision remains open.

@@

    \includegraphics[width=0.980\linewidth]{figures/izhikevich.pdf}%png}% https://www.overleaf.com/5625872443qpcwrkssgbsf

@@

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/conv_HD-SNN.pdf}
    % \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \caption{
    % \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \textbf{Applying HD-SNN to the task of motion detection.} \textit{(Left)}~We plot here as a raster plot a slice of the input event stream (showing ON spikes in red and OFF spikes in blue for each vertical position and time for a given horizontal position at the center of the sensor), with a pure translation motion upwards. A spatio-temporal convolution is applied to the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. The convolution is summed over the two polarities. At any given time (represented by the blue point), we show the extent of the kernels (green dashed frame). \textit{(Middle)}~There are as many kernels as there are possible motion classes in the recording, and we show the activation resulting from the convolution for this particular motion. The activation can be above chance (warm colors) or below chance (cold colors). For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the nonlinearity of the MLR model (i.e., the softmax function). The output of the MLR gives a probability for each class associated with a particular kernel. \textit{(Right)}~By adding a spiking mechanism, here a winner-takes-all associated to a thresholding, we obtain as output of the HD-SNN model a new spike train with the different spikes associated to a specific motion class (here red is associated to an upward motion). Most of the motion is detected accurately. }
    \label{fig:model}
\end{figure}

@@

By pruning weights, we observe that the evolution of accuracy as a function of the log percentage of active weights fits well a sigmoid curve. Half-saturation level is reached at about $3.5\times 10^{-3}\%$ of active weights, corresponding in our setting to a total amount of $6$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained to its top performances when dividing the number of computations by a factor up to about $200$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, thus demonstrating the efficiency of the presented method.

%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/accuracy.pdf}
    \caption{Accuracy as a function of the number of computation load for the model with heterogeneous delays (blue line) and for a method using 2D time surfaces (red dot)~\citep{grimaldi_robust_2022}. The relative computational load (on a log axis) is controlled by changing the percentage of active weights relative to the dense convolution kernel. We observe a similar accuracy than HOTS, yet that our model could achieve a similar accuracy with significantly fewer coefficients.}
    \label{fig:accuracy}
\end{figure}

% \note{try also a quantization of the weights (binary?)}
%
\section{Discussion}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN with heterogeneous delays and shown how it compares favorably to a state-of-the-art event-based classification algorithm for a visual motion detection task. The learned model shares a number of similarities with neurobiological anatomical observations, as well as with behavioral results. The event-based computations of our method can be drastically reduced by pruning synapses, while maintaining top classification performance. This shows that we can take advantage of the precise timing of spikes to improve the performance of neural computations.
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
The HD-SNN model that has been trained and evaluated on a complex motion detection task. The model has been defined to provide optimal detection of event-driven spatio-temporal motifs. We have shown that the model, when trained on a dataset of natural images with realistic eye movements, learns kernels similar to those found in the early visual cortex~\cite{kremkow_push-pull_2016}. %We then demonstrated that the model shows similarities to biological responses.
We have evaluated the computational cost of this model when implemented in a setting similar to event-based hardware. We show that the use of heterogeneous delays may be an efficient computational solution for future neuromorphic hardware, but also a key to understanding why spikes are a universal component of neural information processing.

@@

In conclusion, the model that we have presented provides a way to process event-based signals in an efficient manner. We have shown that we can train the model semi-supervised, knowing \emph{what} output label, but not \emph{when} it occurs. Another perspective would be to extend the model to a fully self-supervised learning paradigm, i.e., without any labeled data~\citep{barlow_unsupervised_1989}. This type of learning is thought to be prevalent in the central nervous system and, assuming the signal is sparse~\citep{olshausen_emergence_1996}, one could extend these Hebbian sparse learning schemes to spikes~\citep{perrinet_emergence_2004, masquelier_competitive_2009}. We expect that this would be particularly useful for exploring neurobiological data. Indeed, there is a large literature showing that brain dynamics often organize into stereotyped sequences, such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007}, or hippocampal sequences~\citep{pastalkova_internally_2008, villette_internally_2015}. These patterns are stereotyped and robust, as they can be activated in the same pattern from day to day~\citep{haimerl_internal_2019}. In contrast to conventional methods of processing neurobiological data, such an event-based model would be able to answer key questions about the representation of information in neurobiological data, and it would open up possibilities in the field of computational neuroscience. Furthermore, it would open up possibilities in the field of machine learning, especially in computer vision, to address current key concerns such as robustness to attacks, scalability, interpretability, or energy consumption.
%
%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
%
\bmhead{Acknowledgments}
%
% Acknowledgement hidden for review.
\Acknowledgments
%
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Conflict of interest}
Not applicable.
% The authors do not have any financial or non-financial interests directly or indirectly related to this work.

>>>>

This research was funded by the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{APROVIS3D}{http://aprovis3d.eu/}''). 

@@

and from A*Midex grant number AMX-21-RID-025
``\href{Polychronies}{https://laurentperrinet.github.io/grant/polychronies/}''. 
\newcommand{\Acknowledgments}{
Thanks to % Thomas Schatz, Rosa Cossart,
Hugo Ladret, Camille Besnainou and Jean-Nicolas Jérémie for useful discussions prior to the elaboration of this work. 
%For the purpose of open access, the author has applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission. This research was funded, in whole or in part, by [Organisation name, Grant #]. 
A CC-BY public copyright license has been applied by the authors to the present document and will be applied to all subsequent versions up to the Author Accepted Manuscript arising from this submission, in accordance with the grant’s open access conditions. 
\newcommand{\DataAvailability}{
This works is made reproducible using the following tools. First the code reproducing all figures is available on~\href{ https://github.com/SpikeAI/pyTERtorch}{GitHub}. % todo: open-up the repo
Find also the associated~\href{https://www.zotero.org/groups/4776796/fastmotiondetection}{zotero group} used to gather relevant literature on the subject.

@@

\jyear{2023}%
%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}[theorem]{Proposition}% 
%
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

@@

@@

@@

@@

\abstract{
\Abstract

\keywords{\Keywords}
%
\maketitle
%
\section{Introduction}
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable property of being able to react at any time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation both use. Artificial convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as tensors, storing visual information densely across the visual topography, with different properties represented in different channels. These networks are known to mimic many properties of biological systems, such that each can be attributed a ``Brain Score''~\citep{schrimpf_brain-score_2020}, yet this score takes no account of inference speed or energy consumption. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. However, they do not take advantage of the dynamics inherent in the way we perceive our natural environment. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. The third generation of artificial neural networks, known as spiking neural networks (SNNs), incorporates this temporal dimension into the way they perform their computations. They are of interest for computational neuroscience because they provide a better model of the biological neuron. Some approaches have developed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses accurate timing of spikes attempts to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been recently extended. This method uses a surrogate gradient and is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of firing-rate based networks. The question of the advantage of using spikes in machine learning and computer vision remains open.

@@

    \includegraphics[width=0.980\linewidth]{figures/HDSNN_graph.pdf}%png}% https://www.overleaf.com/5625872443qpcwrkssgbsf

@@

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/HDSNN_conv.pdf}
    % \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \caption{
    % \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \textbf{Applying HD-SNN to the task of motion detection.} \textit{(Left)}~We plot here as a raster plot a slice of the input event stream (showing ON spikes in red and OFF spikes in blue for each vertical position and time for a given horizontal position at the center of the sensor), with a pure translation motion upwards. A spatio-temporal convolution is applied to the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. The convolution is summed over the two polarities. At any given time (represented by the blue point), we show the extent of the kernels (green dashed frame). \textit{(Middle)}~There are as many kernels as there are possible motion classes in the recording, and we show the activation resulting from the convolution for this particular motion. The activation can be above chance (warm colors) or below chance (cold colors). For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the nonlinearity of the MLR model (i.e., the softmax function). The output of the MLR gives a probability for each class associated with a particular kernel. \textit{(Right)}~By adding a spiking mechanism, here a winner-takes-all associated to a thresholding, we obtain as output of the HD-SNN model a new spike train with the different spikes associated to a specific motion class (here red is associated to an upward motion). Most of the motion is detected accurately. }
    \label{fig:model}
\end{figure}

@@

By pruning weights, we observe that the evolution of accuracy as a function of the log percentage of active weights fits well a sigmoid curve. Half-saturation level is reached at about $3.5\times 10^{-3}\%$ of active weights, corresponding in our setting to a total amount of $6$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained to its top performances when dividing the number of computations by a factor up to about $200$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, thus demonstrating the efficiency of the presented method. 
%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}%[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/accuracy.pdf}
    \caption{Accuracy as a function of the number of computation load for the model with heterogeneous delays (blue line) and for a method using 2D time surfaces (red dot)~\citep{grimaldi_robust_2022}. The relative computational load (on a log axis) is controlled by changing the percentage of active weights relative to the dense convolution kernel. We observe a similar accuracy than HOTS, yet that our model could achieve a similar accuracy with significantly fewer coefficients.}
    \label{fig:accuracy}
\end{figure}
%
\section{Discussion}
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN with heterogeneous delays and shown how it compares favorably to a state-of-the-art event-based classification algorithm for a visual motion detection task. The learned model shares a number of similarities with neurobiological anatomical observations, as well as with behavioral results. The event-based computations of our method can be drastically reduced by pruning synapses, while maintaining top classification performance. This shows that we can take advantage of the precise timing of spikes to improve the performance of neural computations.
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
The HD-SNN model that has been trained and evaluated on a complex motion detection task. The model has been defined to provide optimal detection of event-driven spatio-temporal motifs. We have shown that the model, when trained on a dataset of natural images with realistic eye movements, learns kernels similar to those found in the early visual cortex~\cite{kremkow_push-pull_2016}. %We then demonstrated that the model shows similarities to biological responses.
We have evaluated the computational cost of this model when implemented in a setting similar to event-based hardware. We show that the use of heterogeneous delays may be an efficient computational solution for future neuromorphic hardware, but also a key to understanding why spikes are a universal component of neural information processing.

@@

In conclusion, the model that we have presented provides a way to process event-based signals in an efficient manner. We have shown that we can train the model semi-supervised, knowing \emph{what} output label, but not \emph{when} it occurs. Another perspective would be to extend the model to a fully self-supervised learning paradigm, i.e., without any labeled data~\citep{barlow_unsupervised_1989}. This type of learning is thought to be prevalent in the central nervous system and, assuming the signal is sparse~\citep{olshausen_emergence_1996}, one could extend these Hebbian sparse learning schemes to spikes~\citep{perrinet_emergence_2004, masquelier_competitive_2009}. We expect that this would be particularly useful for exploring neurobiological data. Indeed, there is a large literature showing that brain dynamics often organize into stereotyped sequences, such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007}, or hippocampal sequences~\citep{pastalkova_internally_2008, villette_internally_2015}. These patterns are stereotyped and robust, as they can be activated in the same pattern from day to day~\citep{haimerl_internal_2019}. In contrast to conventional methods of processing neurobiological data, such an event-based model would be able to answer key questions about the representation of information in neurobiological data, and it would open up possibilities in the field of computational neuroscience. Furthermore, it would open up possibilities in the field of machine learning, especially in computer vision, to address current key concerns such as robustness to attacks, scalability, interpretability, or energy consumption.
%
%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
%
\bmhead{Acknowledgments}
%
\Acknowledgments
%
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Conflict of interest}
Not applicable.
